---
title: "Basic Usage of NLMR"
author: "Marco Sciaini, Craig E.Simpkins, Jakub Nowosad"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: Citations.bib
vignette: >
  %\VignetteIndexEntry{Basic Usage of NLMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, include=FALSE}
library(terra)
library(NLMR)
```

`NLMR` is an R package for simulating neutral landscape models (NLMs): synthetic landscapes used to explore landscape-scale ecological patterns and processes.
The package follows the same general idea as the Python package `NLMpy` [see @Etheringtonnlmr2015], but returns results as `SpatRaster` objects so they can be used directly with `terra` and other raster workflows.
Most of the common NLMs described in the literature are available.
The package also includes tools for classifying continuous outputs into categories.

For more background on neutral landscape models and the package design, see the accompanying publication Sciani, Fritsch, Scherer and Simpkins [-@Sciani2018].

## Quickstart

`NLMR` exposes 15 exported landscape-generation functions covering 16 algorithm variants.
The variants span a spectrum of spatial structure, from spatially random landscapes (`nlm_random`) through intermediate patterned forms to deterministic planar gradients (`nlm_planargradient`) [see @Palmer1992].
The curdling generator, `nlm_curds()`, can produce either random curdling or wheyed curdling, which is why the package has one more algorithm variant than exported function.

The plot below shows a basic random landscape with no intentional spatial structure.

A simple random landscape can be generated like this:

```{r, fig.height=7, fig.width=7, fig.align='center'}
x <- NLMR::nlm_random(ncol = 100, nrow = 100)
terra::plot(x)
```

## Combining landscapes

NLM rasters can be combined to create new landscape patterns.
A base raster can be added to one or more secondary rasters, with optional scaling factors controlling how strongly each secondary raster contributes.
This example combines a gradient with random variation to create a more complex surface.

```{r, fig.height=7, fig.width=7, fig.align='center'}
base <- NLMR::nlm_edgegradient(ncol = 100, nrow = 100)
secondary_1 <- NLMR::nlm_distancegradient(
  ncol = 100,
  nrow = 100,
  origin = c(10, 10, 10, 10)
)
secondary_2 <- NLMR::nlm_random(ncol = 100, nrow = 100)

combined <- base + (secondary_1 + secondary_2 * 0.2)
terra::plot(combined)
```

## Classifying categories

Landscape rasters generated by `NLMR` are continuous by default and often fall between 0 and 1.
Use `landscape_classify()` to convert a continuous raster into categorical classes when you want a discrete landscape representation.
This example turns a continuous gradient into four discrete classes.

```{r fig.height=7, fig.width=7, fig.align='center'}
continuous <- NLMR::nlm_edgegradient(ncol = 100, nrow = 100)
classified <- landscape_classify(continuous, n = 4)
terra::plot(classified)
```

## References
